# Sharing data objects

_Last update: `r system("git log -1 --format=\"%ad (%h)\" -- 03-python_and_r.Rmd", intern = TRUE)`_

## From R to Python, back to R

1. Always load the Python environment with the packages you need.

```{r, decorate=TRUE}
library(reticulate)
use_condaenv("r-python")
```

1.  Load the dataset in R and assign it to an R object. Let's call it `autos`:

```{r, decorate=TRUE}
# R chunk
autos = cars       # assign cars to autos
```

2.  Read the R object from Python by adding the prefix `r.` before the name of the R object `autos`. Then, assign it to a Python object that we will name `autos_py`.

```{python, collapse = TRUE, decorate=TRUE}
# Python chunk
import numpy
import pandas 

autos_py = r.autos    # assign to a Python object
```

3.  Make a calculation between two columns in the dataset (distance and speed), and assign it to a new column in the dataset with `autos_py['time']`.

```{python, decorate=TRUE}
# Python chunk
autos_py['time'] = autos_py['dist'] / autos_py['speed']   # calculate on variables
```

4.  From R, read the Python object `py$autos_py` and plot the dataset with the new column `time`, that you obtained in Python. Observe that we added the prefix `py$` in front of the Python object `autos_py`:

```{r, decorate=TRUE, fig.asp=1}
# R chunk
plot(py$autos_py)          # plot a Python data object
```

## From Python to R, back to Python


